---
title: "evolvis Summary"
author: "Stanley J Johnson, Kyle S Johnston"
date: "2015-03-18"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
**STAT290 Final Project**

Kyle S Johnston (SUNet ID kjohnsto)

Stanley J Johnson (SUNet ID sjj)

## Introduction 

**evolvis** is **evol**ution **vis**ualization

The goal of evolvis is a generalized toolset for generating a 
lineage of the components of some composition, such as a document, 
from a series of events such as revisions, and visualizing the lineage 
interactively.

The evolvis package provides three categories of functions
which work together:

- XML transform
- Text differencing
- Interactive visualization

### XML transform
Transforms XML documents into a series of text revisions.  For the 
initial use case these are WikiMedia documents, which is the XML
schema for Wikipedia pages.  When connected to the internet, 
XML documents can be downloaded on demand.

### Text differencing
The `evolution` function is the main entry point, and returns an
object of type `evolution`. It uses formula syntax to express the 
evolution of differences in the response variable `y` as a function 
of the version `v(...)` by content class, 
for example `texts ~ v(revisionNum) + author`. The parameter 
`diff.fun` is required and specifies the differencing function.

### Interactive visualization
The `plot.evolution` function is an S3 dispatch method for the
internal generic `plot` function.  It requires the `ggvis`
package and produces a `ggvis` object which can be rendered 
using the generic `print` method or `ggvis::view_dynamic` 
or `ggvis::view_static`.

### Background
The package was inspired by the IBM Research *History Flow* project
https://www.research.ibm.com/visual/projects/history_flow/
for visualizing the history and composition of Wikipedia articles.

## Summary of current work
The fundamental structure of the package is complete and can be 
demonstrated, but we consider it a work in progress since we've learned 
some things along the way.  We started with the IBM Research 
History Flow visualization project as a baseline concept. With the 
encouragement of Professor Naras we sought in this project to expand on 
the concepts of History Flow as follows: 

### Project Proposal

The following is from the text of the project proposal submitted.

**1. Generalize the IBM History Flow design** 

Show change in the composition of Wikipedia articles as well as other types of objects over time. Three use cases will be implemented for the first release. Two of them are noted as “with vignette” to be included in the package for illustration:

- 1.1. Visualize Wikipedia article history by passing a Wikipedia URL

- 1.2. Visualize Wikipedia article history by passing a character the full history XML document obtained from the Wikimedia API (with vignette)

- 1.3. Visualize the history of an investment portfolio composition by security and asset class (with vignette)

**2. Extend graphical and presentation options**

The following visualization capabilities will be implemented:

- 2.1. Implement visualizations in a Web UI based on the ggvis package

- 2.2. Use alpha/opacity so that components of content can overlap in the timeline and look good

- 2.3. Allow user-defined y-axis relative position (stacking) of each visual component

### Current Progress

**Notable features implemented**

- Implements the Longest Common Substring algorithm in function `lcs` as described in  http://en.wikipedia.org/wiki/Longest_common_subsequence_problem.  This algorithm is the engine under the implemented `testDiff` differencing function.

- Implements a `textDiff` differencing algorithm informed by the approach described in  http://en.wikipedia.org/wiki/Diff_utility.

- The model formula syntax `y ~ x + x ...` is implemented by the `evolution` function for intuitive and flexible specification of the differencing task to be performed.

- Uses `XML` SAX stream parsing to extract text revisions and metadata from the WikiMedia XML schema.

- Implements S3 method dispatch in the function `plot.evolution`.

- Leverages the `ggvis` package to construct the visualization as an object extensible via `ggvis`.

- Visualization uses model formula `terms` from the `evolution` object to construct and label the visualization.

- Visualization renders mouse-over tooltips in interactive mode so that evolution content can be inspected interactively.

**Future work**

We believe the design forms a solid foundation for the intended generalization of evolution visualization, as well as some related utilities like differencing.  Our focus on a flexible and usable foundation probably prevented us from fully meeting the goals of this proposal in the "version 0.1" Final Project timeframe, but on the other hand makes the package more viable going forward.

With that in mind, these are some immediate areas for future work:

- Implement a `wordDiff` differencing algorithm for differencing with a *word* as the smallest token.  This seems appropriate for a prose document.  By contrast, the current `textDiff` implementation in evolvis does *character* token differencing, and a typical code diff utility does *line* token differencing.

- Implement date-time support for terms in the `evolution` formula and object.  This will allow for the timelined plotting that is supported by `plot.evolution`.

- Consider other differencing algorithms such as numeric or binary data types.

- Add more options and interactivity to the visualization.

## Example

**A minimal example demonstrating differencing and visualization**

Create toy data: 
```{r}
library(evolvis)

text1 <- "This is a short sentence."
text2 <- "This is a very short sentence."
text3 <- "This is a very, very short sentence."
texts <- c(text1, text2, text3)
revisionNum <- c(1,2,3)
ThreeAuthors <- c("Larry", "Curly", "Moe")
component <- rep("Body", times=3)
testData <- data.frame(texts, revisionNum, ThreeAuthors, component,
                          stringsAsFactors=FALSE)
```

Create the `evolution` object 
```{r}
testEvo <- evolution(texts ~ v(revisionNum) + ThreeAuthors, 
                         data=testData, diff.fun=textDiff)
```

Create and view the visualization `ggvis` object all at once 
```{r}
vis <- plot(testEvo)
vis
```

Explicit call, plus more options for the plot method
```{r}
vis <- evolvis:::plot.evolution(testEvo, 
                                startAtTop = FALSE, 
                                age = FALSE, 
                                xtitle = "Timeline")
vis
```

Specify the port then use an external browser at `http://localhost:3838`
```{r}
# ggvis::view_dynamic(vis, port = 3838)
```

